---
title: "Module_2_ ARROW Pipes"
author: "Elyse, Jeanne & Sheila"
format: 
  html: 
    toc: true
    toc-depth: 4
    toc-location: right
    code-fold: true
  pdf:
    toc: true
    toc-depth: 4
    toc-location: right
editor: visual
---

## 🎯 Learning Objectives (Module 2)

By the end of this module, you will be able to:

-   **Understand** the mindset shift from "loading" to "viewing" data

-   **Read and explore** large datasets without loading them into memory

-   **Convert** between file formats (CSV ↔ Parquet) efficiently

-   **Apply** basic dplyr operations with Arrow's lazy evaluation

-   **Optimize** query performance through proper `collect()` placement

-   **Build** multi-step pipelines that process gigabytes of data

## 🧠 The Mindset Shift: "View" Don't "Load"

### 💡 Traditional R Approach vs. Arrow Approach

**Traditional R thinking**:

```         
# Load EVERYTHING into memory first 

data <- read_csv("huge_file.csv")  # 💥 Crash! 

data |> 
  filter(year == 2023)
```

**Arrow thinking**:

```         
# Create a "view" of the data, then filter 

data <- open_dataset("huge_file.csv")  # ✅ Instant! d

ata |> filter(year == 2023) 
  |> collect()  # Only brings filtered results to memory
```

## 💡 **Rule of thumb:**

-   **\< 100MB**: Regular R is fine

-   **100MB - 1GB**: Arrow is nice to have

-   **1GB+**: Arrow becomes really valuable

-   **RAM size+**: Arrow is essential

### 🔍 Key Concept: Lazy Evaluation

Arrow uses **lazy evaluation** - it builds up a query plan without actually executing it until you call `collect()`.

Think of it like:

-   **Traditional R**: "Cook the entire meal, then throw away what you don't want"

-   **Arrow**: "Plan the meal, shop for only what you need, then cook just that"

## 🚀 Arrow Basics (10 minutes)

### Setting Up Our Big Data Playground - done in pre-workshop materials

We're working with the **real Seattle Library dataset** - over 40 million rows of checkout data!

### Step 1: Create a Directory

First, let's create a special folder to store our data:

```{r eval=FALSE}

# Create a "data" directory if it doesn't exist already 
# Using showWarnings = FALSE to suppress warning if directory already exists  

dir.create("data", showWarnings = FALSE)
```

### Step 2: Download the Dataset

Now for the fun part! We'll download the Seattle Library dataset (9GB).

⚠️ **Important:** This is a 9GB file, so:

-   Make sure you have enough disk space

```{r eval=FALSE}

# Download Seattle library checkout dataset: 

# 1. Fetch data from AWS S3 bucket URL 
# 2. Save to local data directory 
# 3. Use resume = TRUE to allow continuing interrupted downloads  

curl::multi_download("https://r4ds.s3.us-west-2.amazonaws.com/seattle-library-checkouts.csv", "data/seattle-library-checkouts.csv", resume = TRUE )
```

Why USE: [`curl::multi_download()`](https://jeroen.r-universe.dev/curl/reference/multi_download.html)

-   Shows a progress bar (great for tracking large downloads)
-   Can resume if interrupted (super helpful for big files!)
-   More reliable than base R download methods

### Step 3: Verify the Download

After the download completes, let's make sure everything worked:

```{r}
# Check if the Seattle library dataset file exists and print its size:
# 1. Verify file exists at specified path
# 2. Calculate file size in gigabytes by dividing bytes by 1024^3

file.exists("data/seattle-library-checkouts.csv")
file.size("data/seattle-library-checkouts.csv") / 1024^3  # Size in GB
```

"Who thinks their computer could handle loading 9GB into memory?"

### 🔬 The `open_dataset()` Magic

Now let's see the fundamental difference between `read_csv()` and `open_dataset()`:

DON'T RUN!

```{r eval=FALSE}

# Traditional approach - would crash most computers!

#seattle_library_checkouts <- read_csv("data/seattle-library-checkouts.csv") # DON'T RUN! 
```

RUN

```{r}

#load in the packages and install if needed with code below


# Function to check and install required packages
required_packages <- c("tidyverse", "arrow")

# Install missing packages
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
    #library(pkg, character.only = TRUE)
  }
}

#Load libraries
lapply(required_packages, library, character.only = TRUE)
```

## What `open_dataset()` does:

-   Creates an Arrow dataset object that "points to" your CSV file

-   Doesn't actually load the data into memory yet

-   Acts like a "view" or "window" into your data

-   When you run this code with [`open_dataset()`](https://arrow.apache.org/docs/r/reference/open_dataset.html), Arrow does something clever:

    1.  It peeks at the first few thousand rows

    2.  Figures out what kind of data is in each column

    3.  Creates a roadmap of the data

    4.  Then... it stops!

    That's right - it doesn't load the whole 9GB file. Imagine Arrow as a really efficient librarian who:

    -   Creates an index of where everything is

    -   Only gets books (data) when you specifically ask for them

    -   Keeps track of what's where without moving everything

```{r}
# Arrow approach - creates a "view" without loading 
seattle_csv <- open_dataset("data/seattle-library-checkouts.csv", format = "csv")

# View Object
seattle_csv

library(glue) # string interpolation - cleaner alternative to paste()

# Check out how much memory this is using.
glue("Memory used by Arrow object: {format(object.size(seattle_csv), units = 'KB')}")

# Let's see what the file size we are actually working with 
file_size_bytes <- file.size("data/seattle-library-checkouts.csv")
file_size_gb <- file_size_bytes / (1024^3)  # Convert to GB
glue("Estimated file size: {round(file_size_gb, 2)} GB")

# Peek at the structure without loading 
seattle_csv |> 
  glimpse()
```

## 🔍 **Handling the ISBN Null Issue**

When we glimpse our dataset, we notice that the **ISBN column is showing as null**. This is a common issue when working with large datasets where Arrow's automatic schema detection encounters parsing challenges.

### Why This Matters

**For Arrow:**

-    Arrow creates a schema based on the first few thousand rows it samples

-    If it encounters mixed data types or parsing issues in the ISBN field, it may default to null

-    This means we lose valuable bibliographic information that could be used for:

    -    Identifying unique editions of books

    -    Linking to external book databases

    -    Analyzing publication patterns

```{r}
# Remove the current object that has the null ISBN issue
rm(seattle_csv)

# Recreate Seattle library checkout.csv data:
# 1. Specify the CSV file path
# 2. Set ISBN column to be read as string type to preserve leading zeros
# 3. Define CSV as the file format
seattle_csv <- open_dataset(
  sources = "data/seattle-library-checkouts.csv", 
  col_types = schema(ISBN = string()),
  format = "csv"
)

#inspect
glimpse(seattle_csv)
```


**Key insight**: Arrow creates a "catalog" or "view" of your data without actually reading it into memory!

### 🎯 **Understanding `collect()`: The Bridge Between Arrow and R**

**What is `collect()`?**

-   `collect()` is the function that **executes** your Arrow query and brings the results into regular R memory

-   Think of it as the "Go!" button that turns your query plan into actual data

### 🔍 Exploring Structure Without Loading

```{r}

# 1. BASIC DATASET INFO (no data loading)
glue("=== DATASET OVERVIEW ===")
glue("File size: {format(file.size('data/seattle-library-checkouts.csv'), units = 'GB')}")
glue("Memory used by Arrow object: {format(object.size(seattle_csv), units = 'KB')}")

# 2. SCHEMA EXPLORATION (metadata only)
seattle_csv$schema  # Column types
glue("Number of columns: {ncol(seattle_csv)}")

# 3. SMART SAMPLING (minimal data loading)
seattle_csv |> 
  slice_sample(n = 1000) |>  # Random sample instead of just head()
  glimpse()

# 4. COLUMN-WISE EXPLORATION (targeted queries)
seattle_csv |> 
  summarise(
    total_rows = n(),
    across(where(is.character), ~n_distinct(.x, na.rm = TRUE)),
    across(where(is.numeric), list(min = ~min(.x, na.rm = TRUE), 
                                   max = ~max(.x, na.rm = TRUE)))
  ) |> 
  collect()

##instead save so you don't have to re-processing the massive dataset to explore later!:

# Save the summary (this might take a few minutes for 10GB)
data_summary <- seattle_csv |> 
  summarise(
    total_rows = n(),
    across(where(is.character), ~n_distinct(.x, na.rm = TRUE)),
    across(where(is.numeric), list(min = ~min(.x, na.rm = TRUE), 
                                   max = ~max(.x, na.rm = TRUE)))
  ) |> 
  collect()

# Then print when it's done
data_summary

# 5. CHECK FOR MISSING DATA
missing_nas <- seattle_csv |> 
  summarise(across(everything(), ~sum(is.na(.x)))) |> 
  collect()

#inspect
missing_nas

# Check for missing values in character columns
missing_char <- seattle_csv |> 
  summarise(across(where(is.character), ~sum(is.na(.x) | .x == ""))) |> 
  collect()

#inspect
missing_char

```

```{r}
# Get a cleaner view with percentages
total_rows <- seattle_csv |> 
  summarise(n = n()) |> 
  collect() |> 
  pull(n)

missing_char |> 
  pivot_longer(everything(), 
               names_to = "column", 
               values_to = "missing_count") |> 
  mutate(missing_percentage = round(missing_count / total_rows * 100, 2)) |> 
  arrange(desc(missing_count))
```

### 📋 "Look Ma, No Loading!"

Showing Query Plans with `show_query()`

Before we execute anything, let's see what Arrow plans to do:

```{r eval =FALSE}
query_plan <- seattle_csv |>   
  filter(CheckoutYear == 2020) |>   
  group_by(MaterialType) |>   
  summarise(total_checkouts = sum(Checkouts))  # See what Arrow will do (without doing it!)

query_plan |> 
  show_query()  # This shows the execution plan - Arrow is incredibly smart about optimization!
```

## Reading the Plan (bottom to top):

**Node 0: SourceNode{}**

-   Start with your CSV file

**Node 1: FilterNode{filter=(CheckoutYear == 2020)}**

-   Filter to only 2020 data (reduces data early!)

**Node 2: ProjectNode{projection=\["total_checkouts": Checkouts, MaterialType\]}**

-   Select only the columns needed: `Checkouts` (renamed to `total_checkouts`) and `MaterialType`

**Node 3: GroupByNode{keys=\["MaterialType"\], aggregates=\[hash_sum(...)\]}**

-   Group by MaterialType and sum up the checkouts

**Node 4: SinkNode{}**

-   Final output destination

## What this shows you:

✅ **Arrow is being smart!**

-   It filters first (reduces data volume)

-   Only selects needed columns (reduces memory)

-   Does the grouping efficiently with hash operations

-   Plans the whole operation before executing

**When do you need `collect()`?**

**The Golden Rule:**

-   **Arrow operations**: filter, select, group_by, summarise → **no collect() needed**

-   **R operations**: ggplot, view(), head(), mathematical operations → **collect() first**

### 🤔 **Why This Matters**

Without `collect()`, you're working with an Arrow query object:

```{r}
# This creates a PLAN, not data
query <- seattle_csv |> 
  filter(CheckoutYear == 2023)

class(query)  # "ArrowTabular" - not a data frame!

# This executes the plan and gives you data  
actual_data <- query |> 
  collect()
class(actual_data)  # "data.frame" - now you can use it in R!
```

## 🏗️ File Format Magic (15 minutes)

### 📊 CSV vs Parquet: The Speed Revolution

Let's convert our massive CSV to Parquet and see the dramatic improvements:

```{r}
# Convert the entire dataset efficiently 

seattle_csv |>   
  write_dataset("data/parquet/seattle_parquet/", format = "parquet")
```

```{r eval=FALSE}
# Compare file sizes
csv_size <- file.size("data/seattle-library-checkouts.csv") 
parquet_dir_size <- sum(file.size(list.files("data/parquet/seattle_parquet/",                                              
                                             full.names = TRUE, recursive = TRUE))) 
# Show the comparison
glue("Original CSV size: {csv_size}")
glue("Parquet size: {parquet_dir_size}")  # Changed from parquet_size
glue("Size reduction: {round(as.numeric(csv_size) / as.numeric(parquet_dir_size), 1)}x smaller!")
```

**🎯 Best practice for reproducible analysis:**

1.  Share your **code** on GitHub (always \< 1MB)

2.  Include **sample data** for testing (\< 25MB)

3.  **Document** where to get full dataset

4.  **Provide** conversion scripts for creating Parquet files

### ⚡ Speed Test: CSV vs Parquet

```{r}
# Time reading and filtering
csv_time <- system.time({csv_result <- seattle_csv |>      
  filter(CheckoutYear == 2020) |>      
  collect() })

seattle_parquet <- open_dataset("data/parquet/seattle_parquet/")

parquet_time <- system.time({parquet_result <- seattle_parquet |>      
  filter(CheckoutYear == 2020) |>      
  collect() })  

#check the speed of files

glue("CSV time: {round(csv_time[3], 2)} seconds")
glue("Parquet time: {round(parquet_time[3], 2)} seconds")

if (parquet_time[3] > 0) {
  glue("Speed improvement: {round(csv_time[3] / parquet_time[3], 1)}x faster")
} 
 
# Verify we got the same results 

glue("CSV result rows: {nrow(csv_result)}")
glue("Parquet result rows: {nrow(parquet_result)}")
```

### 🔄 Working with Your Optimized Dataset

```{r}

#check out glimpse
seattle_parquet |>    
  glimpse()  

# Let's see what columns we have to work with 
seattle_parquet |>    
  head() |>    
  collect()

# Quick exploration - much faster now! 
```

**Key takeaways**:

-   Parquet files are typically 3-5x smaller than CSV

-   Parquet files are often 5-10x faster to read

-   Parquet preserves data types (no more parsing!)

## 🛠️ Your First Big Data Pipeline (20 minutes)

Now let's build progressively more complex pipelines, always remembering: **filter early, collect late!**

### 🎯 Exercise 1: Basic Filtering (5 minutes)

**Question**: "What happened to library usage during the pandemic year of 2020?"

```{r eval=FALSE}
pandemic_data <- seattle_parquet  |>   
  filter(#FILL IN HERE) |>   
  collect()  
# Check what we got - from 40+ million rows to... 
```

```{r}
pandemic_data |>   
  count(MaterialType, sort = TRUE)
```

**Turn and talk** (2 minutes): "What do you notice about the material types during 2020? Any surprises?" Why might we be interested in this year?

**Key insight**: We filtered from 40+ million rows to a manageable subset BEFORE bringing data into memory!

### 📊 Exercise 2: Group and Summarize (7 minutes)

**Question**: "Which types of materials were most popular in 2020 - did people shift to digital?"

#### Build the pipeline: filter → group → summarise → arrange → collect

```{r eval=FALSE}

 material_summary <- seattle_parquet |>   
   filter(#FILL IN HERE) |>   
   group_by(MaterialType) |>   
   summarise(total_checkouts = sum(#FILL IN HERE),     
             avg_checkouts = mean(#FILL IN HERE),     
             checkout_count = n(),     
             .groups = "drop"   ) |>   
   arrange(desc(#FILL IN HERE)) |>   
   collect()

# Look at the summary
material_summary
```

**Turn and talk** (2 minutes): "What do you notice, did people shift to digital 2020? Any surprises?"

Extension to this exercise below

### 📈 Exercise 3: Multi-step Pipeline with Visualization (8 minutes)

**Question**: "How did checkout patterns change month by month during 2020?"

#### Build the pipeline: filter → group → summarise → collect → plot

```{r eval = FALSE}

monthly_trends <- seattle_parquet |>   
  filter(#FILL IN HERE == 2020) |>   
  group_by(#FILL IN HERE, MaterialType) |>   
  summarise(total_checkouts = sum(#FILL IN HERE),.groups = "drop") |>   
  collect() 

#inspect
monthly_trends

```

Create Visualization

```{r}
monthly_trends |>   
  ggplot(aes(x = #FILL IN HERE, y = #FILL IN HERE, color = #FILL IN HERE)) +   
  geom_line(linewidth = 1) +   
  geom_point() +   
  scale_x_continuous(breaks = 1:12, labels = month.abb) +   
  scale_y_continuous(labels = scales::comma) +   
  labs(title = "Seattle Library Checkouts by Material Type (2020)",     
       subtitle = "Monthly trends during the pandemic year - real data!",     
       x = "Month",     
       y = "Total Checkouts",     
       color = "Material Type") +   
  theme_minimal() +   
  theme(legend.position = "bottom")
```

**Gallery walk** (2 minutes): Look at others' results. What patterns do you see?

When was the library closure most visible?

## ⚠️ Troubleshooting Checkpoint (5 minutes)

### 🚨 Common Mistakes and How to Fix Them

#### 1. **Premature `collect()`** - The #1 Error!

```{r}
# ❌ WRONG: Collecting too early 
seattle_parquet |>   
  collect() |>  # Brings ALL data into memory first!   
  filter(CheckoutYear == 2020)   # Then filters - too late!  

# ✅ RIGHT: Filter first, collect last se
seattle_parquet |>   
  filter(CheckoutYear == 2020) |>  # Filter on disk   
  collect() # Only bring filtered data to memory
```

#### 2. **Forgetting `collect()`** - Nothing Happens!

```{r}
# ❌ This creates a query plan but doesn't execute it 
query_only <- seattle_parquet |>
  filter(CheckoutYear==2020) |>
  summarize(total = sum(Checkouts))

print("Query plan created, but no results yet:") 
print(class(query_only))


# ✅ Execute the query with collect() a
actual_result <- 
  query_only |> 
  collect() 
print("Now we have actual results:") 
print(actual_result)

```

#### 3. **File Path Issues**

```{r}
#❌ Common file path errors 

#open_dataset("wrong_path.csv")  # File not found 

#open_dataset("folder_name.csv") # Trying to read folder as file  

# ✅ Check if files exist first 

if (file.exists("data/parquet/seattle_parquet/")) {   
  data <- open_dataset("data/parquet/seattle_parquet/")   
  print("✅ Parquet dataset opened successfully!") 
} else if (file.exists("data/seattle-library-checkouts.csv")) {   
  data <- open_dataset("data/seattle-library-checkouts.csv", format = "csv")   
  print("✅ CSV dataset opened successfully!") 
} else {   
  print("❌ Please check your file path - dataset not found") 
}
```

## 🎯 Arrow Choose-Your-Own Challenge (15 minutes)

Work in pairs. Choose the challenge that matches your comfort level, then try the stretch goals!

### 🟢 Beginner Challenge 1A: Material Type Analysis

**Goal**: Build a dplyr pipeline to summarize checkouts by MaterialType across all years.

**Your mission**:

1.  Start with the Arrow dataset

2.  Group by MaterialType

3.  Calculate total checkouts, average checkouts, and count of records

4.  Sort by total checkouts (highest first)

5.  Don't forget to `collect()`!

**Helpful hints for beginners**:

-   Use `group_by()` to group your data

-   Use `summarise()` to calculate statistics

-   Use `sum()` for totals, `mean()` for averages, `n()` for counts

-   Use `arrange(desc())` to sort from highest to lowest

-   End with `collect()` to bring the results into R

```{r}
# Your code here! Use the real Seattle data
#add to the code below

#material_analysis <- seattle_parquet |>
  # Step 1: Group by MaterialType

  # Step 2: Calculate summaries

  # Step 3: Sort by total checkouts

  # Step 4: Collect the results

# Print your results

```

**Goal**: Analyze checkout trends by year to see library usage patterns.

**Your mission**:

1.  Start with the Arrow dataset

2.  Group by CheckoutYear

3.  Calculate total checkouts and number of unique titles per year

4.  Sort by year (oldest first)

5.  Don't forget to `collect()`!

**Helpful hints for beginners**:

-   Use `group_by(CheckoutYear)` to group by year

-   Use `n_distinct(Title)` to count unique titles

-   Use `arrange(CheckoutYear)` to sort chronologically (no `desc()` needed)

```{r}
# Your code here! Analyze trends over time
#add to the code below

#yearly_analysis <- seattle_parquet |>
  # Step 1: Group by year

  # Step 2: Calculate summaries

  # Step 3: Sort by year

  # Step 4: Collect the results


# Print your results

```

**Stretch goals**:

-   Add a filter for years 2015-2023 only

-   Create a simple bar chart of your results

-   Calculate what percentage each material type represents

### 🟡 Intermediate: Multi-step Pipeline + Visualization

### Intermediate Challenge 1A Extension intermediate

**Goal**: Create a quick visualization to see what is happening

**Your mission**:

1.  Use the 2020 pandemic data we already created

2.  Count how many times each MaterialType appears

3.  Sort from most to least common

4.  Create a horizontal bar chart

5.  Add clear labels and title

```{r}

# Quick bar chart of material types
#add your code below




```

Challenge 1B: Digital vs Physical Comparison

**Goal**: Categorize materials as digital or physical and compare their usage

**Your mission**:

1.  First explore what material types exist in our data

2.  Create logical categories (Digital, Physical, Other)

3.  Group by these new categories

4.  Calculate totals and counts for each category

5.  Sort by total checkouts

```{r}
# See all the material types we have
#add your code below

```

Then categorize them:

```{r}

# Compare digital vs physical
#Add your Code below

```

### Exercise 2 Extension intermediate Challenge 1

**Goal**: Analyze checkout trends by month during 2020 (pandemic year)

**Your mission**:

1.  Filter to 2020 data only

2.  Group by CheckoutMonth

3.  Calculate total checkouts per month

4.  Create a line chart showing the monthly trend

5.  Add a title that mentions "Pandemic Impact" \#

```{r}

#monthly_trends <- seattle_parquet |>
  # Add your pipeline here for monthly 2020 analysis




# Create line chart Here

  

  
```

**Stretch goals**:

-   Add monthly granularity to see seasonal patterns

-   Include confidence intervals or trend lines

-   Compare pre-pandemic (2015-2019) vs. pandemic (2020-2023) trends

### 🔄 Optional Brain Break

Feeling overwhelmed? It's totally normal! Take a brain break - you can always do these laterr:

-   Step away from your computer and stretch

-   Grab some water or coffee

-   Check out the [GitHub examples](https://github.com/your-workshop-repo) for inspiration

-   Ask a neighbor how they're approaching the challenge

## 🧠 Key Takeaways from Module 2

### ✅ What You've Accomplished Today

🎉 **Congratulations!** You've just:

-   Processed datasets without loading them entirely into memory

-   Converted massive files for 10x+ speed improvements

-   Used the same dplyr syntax on much larger datasets

-   Built multi-step analytical pipelines with lazy evaluation

-   Learned to optimize performance with proper `collect()` placement

### 🔑 Essential Arrow Patterns to Remember

1.  **The Lazy Evaluation Pattern**:

    ```         
    open_dataset("file") |>     # Create view 
      filter() |>               # Filter on disk   
      group_by() |>             # Group on disk 
      summarise() |>            # Aggregate on disk 
      collect()                 # Bring results to memory
    ```

2.  **The File Conversion Pattern**:

    ```         
    open_dataset("data.csv") |> 
      write_dataset("data_parquet/", format = "parquet")
    ```

3.  **What's under the hood**

    ```         
    query |> 
      show_query()        # See what will happen 

    query |> 
      collect()           # Actually execute
    ```

### 🚀 Coming Up: Module 3 - DuckDB Superpowers

In our next module, we'll add even more power to your toolkit:

-   **Complex joins** across multiple datasets

-   **Window functions** for advanced analytics

-   **SQL integration** that feels like dplyr

-   **When to choose** Arrow vs. DuckDB for different tasks

**The promise continues**: Same familiar syntax, even more analytical power!

### 📝 Quick Self-Assessment

Before we move on, take 30 seconds to reflect:

**I feel confident that I can**:

-   Explain the difference between "loading" and "viewing" data

-   Use `open_dataset()` instead of `read_csv()` for large files

-   Place `collect()` in the right spot in my pipelines

-   Convert CSV files to Parquet format

-   Build multi-step Arrow pipelines

**Please write on a sticky note if you'd like more practice with**:

-   Understanding when to use `collect()`

-   File format conversions

-   Complex pipeline building

-   Performance optimization

-   Troubleshooting errors

*Ready to add database superpowers to your toolkit? Let's continue to Module 3: DuckDB!*

*NSF Acknowledgement: This material is based upon work supported by the National Science Foundation under Grant #DGE-2222148. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the National Science Foundation.*
