---
title: "PRE-Workshop Data Download file"
format: 
  pdf:
    toc: true
    toc-depth: 4
    toc-location: right
editor: visual
---

Welcome to the "*R You out of Memory Short Course*" at the 2026 useR!! confernce. We are so glad you will be joining us. In preperation for a productive learning session we have some pre work for you to do.

As a REMINDER: When you open this Quarto document (.qmd file):

1. **\*\*Make sure it's in your project folder\*\*** alongside your data folder and .Rproj file
2. **\*\*If it's not in the right place:\*\***
 - Use **\*\*File → Save As\*\*** and navigate to your \`useR2026_bigdata_shortcourse\` folder
 - Or drag/move the file into your project folder using your computer's file manager
3. **\*\*Open your .Rproj file\*\*** to ensure RStudio is working in the correct directory
4. **\*\*Verify everything is ready:\*\***

Our dataset is to large to be housed in github or as a file. We can add it as a zip file but after all we are laearning how to havndle large data in R So...we will do so.

The dataset we are using will take anywhere for 8 minutes -15 minutes to download.

We want you to set up a Project folder and download it to their so lets get doing that.

🤔 Data Scientist Thinking: When working with big data, we need to consider both computational efficiency and memory management. How might these features impact our daily workflow?

Resource: [Apache cookbook](https://arrow.apache.org/cookbook/r/index.html)

```{r}
#Load Library
library(tidyverse)
```

## Downloading the Data

A dataset of item checkouts from Seattle public libraries, available online at [data.seattle.gov/Community/Checkouts-by-Title/tmmm-ytt6](https://data.seattle.gov/Community/Checkouts-by-Title/tmmm-ytt6).

### Step 1: Create a Directory

First, let's create a special folder to store our data:

```{r eval=FALSE}
# Create a "data" directory if it doesn't exist already
# Using showWarnings = FALSE to suppress warning if directory already exists

dir.create("data", showWarnings = FALSE)
```

### Step 2: Download the Dataset

Now for the fun part! We'll download the Seattle Library dataset (9GB).

⚠️ **Important:** This is a 9GB file, so:

-   Make sure you have enough disk space

```{r eval=FALSE}
# Download Seattle library checkout dataset:
# 1. Fetch data from AWS S3 bucket URL
# 2. Save to local data directory
# 3. Use resume = TRUE to allow continuing interrupted downloads

curl::multi_download(
  "https://r4ds.s3.us-west-2.amazonaws.com/seattle-library-checkouts.csv",
  "data/seattle-library-checkouts.csv",
  resume = TRUE
)
```

Why USE: [`curl::multi_download()`](https://jeroen.r-universe.dev/curl/reference/multi_download.html)

-   Shows a progress bar (great for tracking large downloads)
-   Can resume if interrupted (super helpful for big files!)
-   More reliable than base R download methods

Resource: [CURL](https://cran.r-project.org/web/packages/curl/vignettes/intro.html)

## 🤔 While the file downloads, let's think about:

1.  Why do we need special tools for such large datasets?

2.  What challenges might we face with traditional R methods?

3.  How might a library use this kind of data?

We will discuss these questions and more when we meet at the short-course.

### Step 3: Verify the Download

After the download completes, let's make sure everything worked:

```{r}
# Check if the Seattle library dataset file exists and print its size:
# 1. Verify file exists at specified path
# 2. Calculate file size in gigabytes by dividing bytes by 1024^3

file.exists("data/seattle-library-checkouts.csv")
file.size("data/seattle-library-checkouts.csv") / 1024^3  # Size in GB
```
